### 2025-06-19 misc/expo/additive.tex
Ben created
update

\edef\lempetridis{\the\thmcount}
\parenproclaim Lemma {\advthm} (Petridis, {\rm 2012}). Let $A$ and $B$ be subsets of an abelian
group and let $K_0$ be such that $|A_0+B| = K_0|A_0|$. Then there exist $A\subseteq A_0$ and
$K\le K_0$ such that $|A+B+C| \le K|A+C|$ for every finite subset $C$ of the group.

\proof We pick $A\subseteq A_0$ nonempty such that the ratio $|A+B|/|A|$, which we shall
set as $K$, is minimised. So $|A+B| = K|A|$ and $|A'+B|\ge K|A'|$ for all $A'\subseteq A_0$.
Now we shall show by induction on $C$ that $|A+B+C|\le K|A+C|$ for all $C$. If $C$ is empty we
are done, of course. Now assume the result holds for $C$ and let $x\notin C$ be arbitrary. We have
$$\bigl| A+(C\cup \{x\}) \bigr| = |A+C| + |A+x| - |A'+x|$$
where $A' = \{a\in A : a+x \in A+C\}$. But adding $x$ to everything in a set does not change its
cardinality, so we have
$$|A+C| + |A+x|-|A'+x| = |A+C| + |A| + |A'|.$$
Also, any element of $A'+B+x$ is an element of $A+B+C$, since $a'+x\in A+C$ for all $a'\in A'$.
Applying the induction hypothesis now gives
$$\eqalign{
\bigl| A + B + (C\cup \{x\}) \bigr| &\le |A+B+C| + |A+B+x| - |A'+B+x| \cr
&\le K|A+B| + K|A| = K|A'| \cr
&= K\bigl| A+(C\cup \{x\})\bigr|,
}$$
which completes the proof.\slug

@@

\proclaim Corollary {\advthm}. If $|A_0 + B|\le K_0|A_0|$, then there is $A\subseteq A_0$ nonempty
and $K\le K_0$ such that $|A+rB| \le K^r |A|$ for all nonnegative integers $r$.

@@

\proof For each $x\in V-W$, pick $v(x)\in V$ and $w(x)\in W$ such that $v(x)-w(x) = x$ (there may
be multiple choices for each of $v(x)$ and $w(x)$, but we arbitrarily pick one).
Define $\phi : U\times (V-W) \to (U-V)\times (U-W)$ by $\phi(u,x) = \bigl(u-v(x), u-w(x)\bigr)$.
If $\bigl(u_1 - v(x_1), u_1 - w(x_2)\bigr) = \bigl(u_2 - v(x_2), u_2 - w(x_2)\bigr)$, then
subtracting the second coordinate of each pair from the first, we find that
$$x_1 = v(x_1) - w(x_1) = v(x_2) - w(x_2) = x_2,$$
and this also implies that $u_1 = u_2$. So $\phi$ gives an injection from $U\times (V-1)$
to $(U-V)\times (U-W)$.\slug

If we define $d(U,V) = |U-V| / \bigl(|U|^{1/2}|V|^{1/2}\bigr)$, then this lemma states
that $d(U,W) \le d(V,U)d(U,W)$. Taking logarithms gives a genuine additive triangle inequality,
but the space of finite subsets of an abelian group does not form a metric space, since
$|U-U|/|U|$ does not equal $1$ in general. In any case, we are now able to prove Ruzsa's
version of Pl\"unnecke's inequality.

@@

>>>>

\edef\lempetridis{\the\thmcount}
\parenproclaim Lemma {\advthm} (Petridis, {\rm 2012}). Let $A_0$ and $B$ be subsets of an abelian
group and let $K_0$ be such that $|A_0+B| = K_0|A_0|$. Then there exist $A\subseteq A_0$ and
$K\le K_0$ such that $|A+B+C| \le K|A+C|$ for every finite subset $C$ of the group.

\proof We pick $A\subseteq A_0$ nonempty such that the ratio $|A+B|/|A|$, which we shall
set as $K$, is minimised. So $|A+B| = K|A|$ and $|A'+B|\ge K|A'|$ for all $A'\subseteq A_0$.
Now we shall show by induction on $C$ that $|A+B+C|\le K|A+C|$ for all $C$. If $C$ is empty we
are done, of course. Now assume the result holds for $C$ and let $x\notin C$ be arbitrary. We have
$$\bigl| A+(C\cup \{x\}) \bigr| = |A+C| + |A+x| - |A'+x|$$
where $A' = \{a\in A : a+x \in A+C\}$. But adding $x$ to everything in a set does not change its
cardinality, so we have
$$|A+C| + |A+x|-|A'+x| = |A+C| + |A| + |A'|.$$
Also, any element of $A'+B+x$ is an element of $A+B+C$, since $a'+x\in A+C$ for all $a'\in A'$.
Applying the induction hypothesis now gives
$$\eqalign{
\bigl| A + B + (C\cup \{x\}) \bigr| &\le |A+B+C| + |A+B+x| - |A'+B+x| \cr
&\le K|A+B| + K|A| - K|A'| \cr
&= K\bigl| A+(C\cup \{x\})\bigr|,
}$$
which completes the proof.\slug

@@

\edef\corpetridis{\the\thmcount}
\proclaim Corollary {\advthm}. If $|A_0 + B|\le K_0|A_0|$, then there is $A\subseteq A_0$ nonempty
and $K\le K_0$ such that $|A+rB| \le K^r |A|$ for all nonnegative integers $r$.

@@

\proof For each $x\in V-W$, pick $v(x)\in V$ and $w(x)\in W$ such that $v(x)-w(x) = x$ (there may
be multiple choices for each of $v(x)$ and $w(x)$, but we arbitrarily pick one).
Define $\phi : U\times (V-W) \to (U-V)\times (U-W)$ by $\phi(u,x) = \bigl(u-v(x), u-w(x)\bigr)$.
If $\bigl(u_1 - v(x_1), u_1 - w(x_2)\bigr) = \bigl(u_2 - v(x_2), u_2 - w(x_2)\bigr)$, then
subtracting the second coordinate of each pair from the first, we find that
$$x_1 = v(x_1) - w(x_1) = v(x_2) - w(x_2) = x_2,$$
and this also implies that $u_1 = u_2$. So $\phi$ gives an injection from $U\times (V-W)$
to $(U-V)\times (U-W)$.\slug

If we define $d(U,V) = |U-V| / \bigl(|U|^{1/2}|V|^{1/2}\bigr)$, then this lemma states:q
that $d(U,W) \le d(V,U)d(U,W)$. Taking logarithms gives a genuine additive triangle inequality,
but the space of finite subsets of an abelian group does not form a metric space, since
$|U-U|/|U|$ does not equal $1$ in general. In any case, we are now able to prove Ruzsa's
version of Pl\"unnecke's inequality.

@@

\medskip\boldlabel Sumsets with distinct summands.
The Pl\"unnecke--Ruzsa inequality can be generalised from iterated sumsets of one fixed set $B$
to apply to sumsets $B_1 + \ldots + B_m$ of different sets. We start with a lemma.

\proclaim Lemma \advthm. Let $G$ be an abelian group and let $A, B_1,\ldots, B_m$
be subsets of $G$ such that for all $1\le i\le m$, $|A+B_i|\le K_i|A|$ for some integer constants
$K_1,\ldots, K_m$. Then there is a subset $A'\subseteq A$ such that
$$|A' + B_1 + \cdots + B_m| \le m^m K_1\cdots K_m |A'|.$$

\proof
Let $e_1,\ldots,e_m$ denote
the standard basis of $\ZZ^m$, set $N_i = K_1\cdots K_m / K_i$ for all $1\le i \le m$,
and let $B_i' = B_i \times \{e_i, 2e_i, \ldots, N_ie_i\} \subseteq G\times \ZZ^m$ for all
$1\le i\le m$. By abuse of notation, consider $A = A\times \{0\}$ to be a subset of $G\times \ZZ^m$
as well, and let $B = \bigcup_{i=1}^m B_i'$. Then
$$|A+B| \le \sum_{i=1}^m |A+B_i'| = \sum_{i=1}^m N_i|A+B_i| \le |A|\sum_{i=1}^m N_i K_i.$$
By Corollary~{\corpetridis} there exists $A'\subseteq A$ such that
$$ |A'+ m B| \le \biggl( \sum_{i=1}^m N_i K_i \biggr)^m |A'|,$$
so we have
$$\eqalign{
(K_1\cdots K_m)^{m-1} |A' + B_1 + \cdots + B_m| &= N_1\cdots N_k |A' + B_1 + \cdots + B_m| \cr
&= |A'+B_1' + \cdots + B_m'| \cr
&\le |A' + mB| \cr
&\le \biggl( \sum_{i=1}^m N_i K_i\biggr)^m |A'| \cr
&\le (m K_1 \cdots K_m )^m |A'|. \cr
}$$
Cancelling out a factor of $(K_1\cdots K_m)^{m-1}$ from both sides proves the result.\slug

If we just want to bound the size of $|B_1+\cdots + B_m|$, we can remove the factor
of $m^m$ on the right-hand side by employing a variant of the ``tensor power trick.''

\proclaim Theorem \advthm. Let $A, B_1, \ldots, B_m$ be subsets of an abelian group $G$ such that
$|A+B_i| \le K_i|A|$ for some integer constants $K_1,\ldots,K_m$. Then
$$|B_1 + \cdots + B_m| \le K_1 \cdots K_m |A|.$$

\proof Let $d$ be a positive integer. We have
$$|A^d + B_i^d| = |A+B_i|^d \le \bigl(K_i |A|\bigr)^d = K_i^d |A^d|$$
for all $1\le i\le m$, so by the previous lemma, there is a subset $A'$ of $A^d$ such that
$$|A' + B_1^d + \cdots + B_m^d| \le m^m (K_1\cdots K_m)^d |A'|
\le m^m (K_1\cdots K_m)^d |A|^d.$$
But this implies that
$$|B_1+\cdots + B_m|^d \le |A' + B_1^d + \cdots + B_m^d| \le m^m (K_1\cdots K_m)^d |A|^d.$$
Taking $d$th roots yields
$$|B_1 + \cdots +B_m| \le m^{m/d} K_1 \cdots K_m |A|,$$
and this holds for all $d$, so by taking $d$ to infinity we can make $m^{m/d}$ as close to $1$ as
we like.\slug

### 2025-06-19 misc/expo/entropy.tex
update

Let $Y$ be the random variable taking the value $0$ if $X^n\in E$ and $1$ if $X^n\in F$. Then
by Corollary~{\corxn}, $\Eta\{X^n\} = n\Eta\{X\}$, and since $X^n$ determines $Y$,
$$\Eta\{X^n\} = \Eta\{Y\} + \Eta\{X^n\given Y\} \ge \Eta\{Y\}.$$
But $\Eta\{Y\} > 0$ for $\eps$ small enough, the normalisation and continuity axioms. So
$\Eta\{X\} \ge \Eta\{Y\}/n > 0$ as well.\slug

>>>>

Let $Y$ be the random variable taking the value $0$ if $X^n\in E$ and $1$ if $X^n\in F$. Then
by Corollary~{\corxn}, $\Eta\{X^n\} = n\Eta\{X\}$, and since $X^n$ determines $Y$,
$$\Eta\{X^n\} = \Eta\{Y\} + \Eta\{X^n\given Y\} \ge \Eta\{Y\}.$$
But $\Eta\{Y\}$ can be taken as close to $1$ as we like by choosing $\eps$ small enough,
by the normalisation and continuity axioms. So
$\Eta\{X\} \ge \Eta\{Y\}/n > 0$.\slug
